---
title: Cancer rates in SC - model diagnostics and visualizations
author: John
date: '2018-09-03'
slug: cancer-rates-in-sc-model-diagnostics-and-visualizations
categories:
  - South Carolina
tags:
  - Bayesian
  - Stan
draft: true
---

# Introduction and purpose

In the [last post](/post/cancer-rates-in-sc-introduction-to-the-tidybayes-package) we scraped new cancer incidence and population data from the South Carolina government website, merged them together, and ran a model to assess relative cancer rates by county. We then visualized the posterior distributions briefly using the `tidybayes` package. Here we dig deeper into model diagnostics using `tidybayes` and also introduce the `loo` package (stands for leave-one-out).

<!--more-->

# Setup and Data

We no longer need the `rvest` and `xml2` packages, so we load the same packages as last time, with the addition of `loo`.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(viridis)
library(stringr)
library(rstanarm)
library(tidybayes)
library(loo)

options(mc.cores = 4)
`%nin%` <- function (x, table) match(x, table, nomatch = 0L) == 0L
```

You can replicate the steps of the last post to get data, but I saved that as a csv file for ease. I also saved the object returned by `stan_glm`.

```{r}
new_cancer_rate <- read_csv("data/cancer_cases_new_sc_2010-2014.xls")
bayes_pois_fit <- load("data/new_cancer_rate_fit.RData")
```

