---
title: Federal R&D Budget (Tidy Tuesday Feb 12, 2019)
author: John
date: '2019-02-16'
slug: federal-r-d-budget-tidy-tuesday-feb-12-2019
categories:
  - General
tags:
  - tidy tuesday
---

# Purpose

The purpose of this is to analyze federal budget data for Tidy Tuesday February 12, 2019. The webpage for the challenge is at the [R2DS website](https://github.com/rfordatascience/tidytuesday/tree/master/data/2019/2019-02-12). The original data source is the [AAAS](https://www.aaas.org/programs/r-d-budget-and-policy/historical-trends-federal-rd).

# Setup

```{r setup, include=FALSE}
library(tidyverse)
```

# Get data

From the website, the code to get the data is as follows:

```{r}
fed_rd <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-02-12/fed_r_d_spending.csv")
# energy_spend <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-02-12/energy_spending.csv")
# climate_spend <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-02-12/climate_spending.csv", col_types = "ccd")
```

There is an issue with the climate spend data, so I downloaded the Excel file from the source site and reimported. I had to rework the cleaning code a bit, mostly to deal with some unfortunate renaming issue in tibble's name repair mechanism (the variable name for the agency gets renamed to `..1`, which confounds the NSE mechanisms of `rename` and other `dplyr` functions). Basically, I fell back to base R's name repair and modified the cleaning code slightly to compensate. However, I don't use the energy and climate dataset below, so I comment those out for now.

One thing that will be interesting: adding in the party of the President. Though Nixon/Ford started in 1969, I'll just start from 1976, when the data begin. This will help with the graphs.

```{r}
party <- tribble(
  ~party,       ~from,  ~to,
  "Republican", 1976,   1977,
  "Democrat",   1977,   1981,
  "Republican", 1981,   1993,
  "Democrat",   1993,   2001,
  "Republican", 2001,   2009,
  "Democrat",   2009,   2017,
  "Republican", 2017,   2019
)
```

Let's add this to the `fed_rd` dataset.

```{r}
fed_rd %>% 
  crossing(party) %>% 
  filter(year >= from & year < to) %>% 
  select(-from, -to) ->
  fed_rd
```



# Exploration

## Federal R&D

The first dataset shows Federal R&D, and looks as follows:

```{r}
head(fed_rd)
```

This already suggests a visualization, so let's try one. The default y-axis label is in dollars, so I use the `scales::dollar_format` function to scale it to billions of dollars, and adjust the label accordingly.

```{r}
ggplot(fed_rd, aes(year, rd_budget, group = department, color = department)) +
  geom_line() +
  geom_rect(aes(xmin = from, xmax = to, ymin = -Inf, ymax = Inf, fill = party), 
            inherit.aes = FALSE, data = party, alpha = 0.2) +
  xlab("Year") +
  ylab("Federal R&D Budget (billions)") +
  scale_color_viridis_d() +
  scale_fill_brewer() +
  scale_y_continuous(labels = scales::dollar_format(scale = 1e-9))
```

It would be interesting to get these budgets as a percentage of discretionary or total outlays. We'll start with discretionary. I use the `scales::percent_format` to label the y-xaxis. The automatic guess from `scales::percent` gives one decimal place, which I thought was silly for this graph, so I put in the extra effort to cut it out.

```{r}
fed_rd %>% 
  mutate(rd_perc_disc = rd_budget / discretionary_outlays) %>% 
  ggplot(aes(year, rd_perc_disc, group = department, color = department)) +
  geom_line() +
  geom_rect(aes(xmin = from, xmax = to, ymin = -Inf, ymax = Inf, fill = party), 
            inherit.aes = FALSE, data = party, alpha = 0.2) +
  xlab("Year") +
  ylab("Federal R&D proportion of discretionary spending") +
  scale_colour_viridis_d() +
  scale_fill_brewer() +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))
```

The same for percentage of total, leaving the extra decimal place in this time because it is informative:

```{r}
fed_rd %>% 
  mutate(rd_perc_tot = rd_budget / total_outlays) %>% 
  ggplot(aes(year, rd_perc_tot, group = department, color = department)) +
  geom_line() +
  geom_rect(aes(xmin = from, xmax = to, ymin = -Inf, ymax = Inf, fill = party), 
            inherit.aes = FALSE, data = party, alpha = 0.2) +
  xlab("Year") +
  ylab("Federal R&D proportion of total spending") +
  scale_colour_viridis_d() +
  scale_fill_brewer() +
  scale_y_continuous(labels = scales::percent) 
```

Ok, so this brief exploration showed that defense spending has dominated other spending, at least as a portion of the federal budget. One thing I did check on at the AAAS site is that this spending is adjusted for inflation. (Probably should have checked that before embarking on this!)

I think it is worth trying to remake the above graph but switching the legend order based on the average spending over time. Here we go:

```{r}
fed_rd %>% 
  mutate(rd_perc_tot = rd_budget / total_outlays,
         department = fct_reorder(department, rd_perc_tot, .fun = mean)) %>% 
  ggplot(aes(year, rd_perc_tot, group = department, color = department)) +
  geom_line() +
  geom_rect(aes(xmin = from, xmax = to, ymin = -Inf, ymax = Inf, fill = party), 
            inherit.aes = FALSE, data = party, alpha = 0.2) +
  xlab("Year") +
  ylab("Federal R&D proportion of total spending") +
  scale_colour_viridis_d(option = "B", direction = -1) +
  scale_fill_brewer() +
  scale_y_continuous(labels = scales::percent) 
```

I think this graph is more informative than the above graph ordered alphabetically by department. I also switched up the color scheme a bit. To me this subtle change made the graph far more informative, with the color legend giving the order of the priority (more or less) of R&D vs the discretionary spending from least to most. This simple change using the `fct_reorder` function from the `forcats` package (loaded as part of `tidyverse`) is very powerful. (I could have also done this in the `ggplot` aesthetic.)

One more quick addition to see if it adds anything:

```{r}
fed_rd %>% 
  mutate(rd_perc_tot = rd_budget / total_outlays,
         department = fct_reorder(department, rd_perc_tot, .fun = mean)) %>% 
  ggplot(aes(year, rd_perc_tot, group = department, color = department)) +
  geom_line() +
  geom_rect(aes(xmin = from, xmax = to, ymin = -Inf, ymax = Inf, fill = party), 
            inherit.aes = FALSE, data = party, alpha = 0.2) +
  geom_smooth(method = "lm", se = FALSE, size = 0.5, linetype = "dotted") +
  xlab("Year") +
  ylab("Federal R&D proportion of total spending") +
  scale_colour_viridis_d(option = "B", direction = -1) +
  scale_fill_brewer() +
  scale_y_continuous(labels = scales::percent) 
```

The fitted lines give a little more information that may help group the R&D budgets even further. We have DOD, in a class by itself (obvious from the raw line plot), HHS and NIH that have a slight trend down over time, DOE and NASA that started higher but have a stronger trend down, then everything else that started low relative to everything else and as a result remain relatively flat.  One final picture, putting the y-axis on log scale:

```{r}
fed_rd %>% 
  mutate(rd_perc_tot = rd_budget / total_outlays,
         department = fct_reorder(department, rd_perc_tot, .fun = mean)) %>% 
  ggplot(aes(year, rd_perc_tot, group = department, color = department)) +
  geom_line() +
  geom_rect(aes(xmin = from, xmax = to, ymin = 0.0001, ymax = Inf, fill = party), 
            inherit.aes = FALSE, data = party, alpha = 0.2) +
  geom_smooth(method = "lm", se = FALSE, size = 0.5, linetype = "dotted") +
  xlab("Year") +
  ylab("Federal R&D proportion of total spending") +
  scale_colour_viridis_d(option = "B", direction = -1) +
  scale_fill_brewer() +
  scale_y_log10(labels = scales::percent) 
```

This graph was produced by switching `scale_y_log10` in for `scale_y_continuous`. You get a whole bunch of junk complaining about how you can't take the log of a negative values (from `geom_smooth`, because the linear trend extends below 0), and `geom_rect` if you keep the `ymin = -Inf`. I went ahead and changed `ymin = 0.01` so we could keep the party shading. The main effect of the log scale was to explore the trends of the "everything else" in the above graphs. What was a relatively flat trend still is, but it is shown that everything in R&D is trending down relative to discretionary spending.

These three graphs all give an interesting picture. Under Republican presidents in general, defense R&D spending has increased, sometimes drastically so, for a few years and started to fall. But as a general trend, defense spending as a percentage of discretionary and total spending has declined since the 1970s, where we were still in the heat of the Cold War. (It's important to remember we're talking about R&D spending.)

# Just for kicks

Ow, my eyes!

```{r}
library(gganimate)
fed_rd %>% 
  mutate(rd_perc_tot = rd_budget / total_outlays,
         department = fct_reorder(department, rd_perc_tot, .fun = mean)) %>% 
  ggplot(aes(x = "", rd_perc_tot, fill = department)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y") +
  scale_fill_viridis_d(option = "B") +
  ggtitle("Federal R&D proportion of total spending") +
  scale_y_continuous(labels = scales::percent) +
  theme_void() +
  transition_time(year) +
  ease_aes("linear")
```

I think I just invented the incredible shrinking pie chart.
  
# Discussion

Honestly, this blog post went in a slightly different direction than I thought. The main takeaway from above is that subtle changes to the coloring, ordering, and scaling of a graph can reveal different insights in the data. I don't think any one of the above is particularly better than any other, just different.

There are a lot of different directions this can go. I thought about using `gganimate` with pie charts just to hurt some eyes. If you were a goverment contractor, you might be interested in forecasts, and it would be easy to apply the techniques from the previous blog posts to do that. I didn't even touch the climate dataset.